NMRlipids IV: Other than PC headgroups

### 2017-03-08 Manuscript/manuscript.tex
Update text

\usepackage{rotating} 
\usepackage{times}
\usepackage{graphicx}
\usepackage{setspace}
\usepackage{amsmath}
\usepackage{epstopdf}
\usepackage[obeyFinal]{easy-todo}
\begin{document}

@@

Based on superficial reading, the conclusions in the literature are roughly
1) glycerol backbone structures are largely similar irrespectively of the headroup \cite{gally81},
2) glycerol backbone and headgroup strucures are similar in model membranes and in bacteria \cite{gally81,??},
3) headgroup structure is phosphatidylserine more rigid than in other phospholipids, which are quite similar to
each others \cite{??}.
In contrast to PC lipids, extensive discussion on glycerol backbone and headgroup structural details do not
exist (as far as I know).

@@

Several simulations containing PE, PG and PS lipids have been published \cite{??},
however, glycerol backbone and headgroup order parameters are not compared to
the experiments (based on superficial reading of literature).

Glycerol backbone and headgroup order parameters for PS lipids from experiments
and simulations are shown in Fig. \ref{HGorderParametersPOPS}.
The preliminary test with CHARMM GUI simulations show suggest that the model
overestimates magnitudes in $\beta$ and g$_3$ order parameters and forking in $\alpha$.
\begin{figure}[]
  \centering
  \includegraphics[width=9.0cm]{../Figs/HGorderparametersPOPS.eps}
  \caption{\label{HGorderParametersPOPS}
    Order parameters for POPS headgroup and glycerol backbone from simulations and experiments without CaCl$_2$ 
    (DOPS from \cite{browning80} contains 0.1M of NaCl, POPC:POPS mixture from \cite{roux90}).
    Absolute values are shown for experimental data, because signs are not known.
    Simulations values are -S$_{CH}$
  }
  \todo{More simulation data for lipids with different headgroups to be collected}
  \todo{CHARMM GUI simulation contains only counter ions as potassium.
    All experiments here contain some amount of sodium salt. The best
    ion concentrations for comparison should be figured out.}
  \todo{Experimental signs of the order parameters would highly useful.}
\end{figure}

Glycerol backbone and headgroup order parameters for PS lipids from experiments
and simulations are shown in Fig. \ref{HGorderParametersPOPS}.
The preliminary test with CHARMM GUI simulations show suggest that the model
\begin{figure}[]
  \centering
  \includegraphics[width=9.0cm]{../Figs/HGorderparametersPOPG.eps}
  \caption{\label{HGorderParametersPOPG}
    Order parameters for PG headgroup and glycerol backbone from simulations and experiments without CaCl$_2$ 
    (POPG from \cite{borle85} contains 10mM of PIPES, DPPG from \cite{wohlgemuth80} contains 10mM PIPES and 100mM CaCl,
    E.Coli PG results from \cite{gally81}).
    Absolute values are shown, because signs are not known for experimental data.
    Glycerol backbone order parameters are negative in simulations, while $alpha$, $\beta$ and
    smaller $\gamma$ are positive.
  }
  \todo{More simulation data for lipids with different headgroups to be collected}
  \todo{CHARMM GUI simulation contains only counter ions as potassium.
    All experiments here contain some amount of sodium salt. The best
    ion concentrations for comparison should be figured out.}
  \todo{Experimental signs of the order parameters would highly useful.}
\end{figure}

@@

Ion binding affinity in lipid bilayer containing PC lipids can be 
measured by using C-H bond order parameters for headgroup $\alpha$ and
$\beta$ carbons by using the electrometer concept. The decrease of these 
order parameters is linearly proportional to the amount of bound charge 
in bilayer \cite{??,catte16}. The electrometer concept can be used also
for bilayers containing PC lipids mixed with charged lipids \cite{macdonald87,roux90,??}.
This is demonstrated in Figs \ref{OrderParametersWithCaCl},
\ref{OrderParametersWithCaClBELOW1M} and \ref{OrderParameterCHANGESWithCaClBELOW1M}, where PC headgroup order parameters 
as a function of CaCl$_2$ concentration are shown from experiments with POPC and PS or PG.
\begin{figure}[]
  \centering
  \includegraphics[width=9.0cm]{../Figs/LIPIDSwithCaCl.eps}
  \caption{\label{OrderParametersWithCaCl}
    PC headgroup order parameters as a function of CaCl concentration from experiments containing charged lipids.
    Pure DPPC data from \cite{akutsu81}, pure POPC data from \cite{altenbach84}, 
    POPC:POPS mixture data from \cite{roux90}, POPC:POPG mixture data with 0.1M NaCl from \cite{macdonald87}
    and POPC:POPG mixture data without NaCl from \cite{borle85}.
  }
  \todo{Check the NaCl concentrations in the samples.}
\end{figure}
\begin{figure}[]
  \centering
  \includegraphics[width=9.0cm]{../Figs/LIPIDSwithCaClBELOW1M.eps}
  \caption{\label{OrderParametersWithCaClBELOW1M}
    Figure \ref{OrderParametersWithCaCl} zoomed to smaller concentrations.
  }
\end{figure}
\begin{figure}[]
  \centering
  \includegraphics[width=9.0cm]{../Figs/CHANGESwithCaCl.eps}
  \caption{\label{OrderParameterCHANGESWithCaClBELOW1M}
    Changes in order parameters
  }
\end{figure}

Order parameters increase when PS or PG are mixed with PC 
in the absence of additional ions, as seen from Figs. 
\ref{OrderParametersWithCaCl} and \ref{OrderParametersWithCaClBELOW1M}.
In electrometer concept this is explained by the tilting of
headgroup more parallel to membrane normal \cite{??}.
The order parameter decrease with increasing CaCl$_2$ concentration 
is explained by the tilting of headgroup more perpendicular
to membrane normal. This decrease is shown to be a good measure
for the amount of bound ions in lipid bilayer with PC lipids \cite{??,catte16}.

The order parameter decrease as a function of added CaCl$_2$
for systems with different amount of negative charge are shown 
in Fig. \ref{OrderParameterCHANGESWithCaClBELOW1M}.
As expected, the order parameter decrease is more 
pronounced with increasing amount of negatively charged lipids in
bilayer. This is explained by the increase of cation concentration
in proximity of bilayer containing negatively charged lipids \cite{??}.

In addition to the changes in PC headgroup order parameters with 
ion binding, also changes in PS and PG headgroup order parameters
are measured.

\begin{figure}[]
  \centering
  \includegraphics[width=9.0cm]{../Figs/PSPGwithCaCl.eps}
  \caption{\label{PSPGchangesWITHCaCl}
    Changes in order parameters
  }
  \todo{Get the small concentration data from the inserts}
\end{figure}

>>>>

\usepackage{rotating} 
\usepackage{times}
\usepackage{graphicx}
\usepackage{setspace}
\usepackage{amsmath}
\usepackage{epstopdf}
\usepackage[obeyFinal]{easy-todo}
\usepackage{csquotes}

\begin{document}

@@

Based on superficial reading, the conclusions in the literature are roughly \\ 
1) glycerol backbone structures are largely similar irrespectively of the headroup \cite{gally81}, \\
2) glycerol backbone and headgroup strucure and behaviour are similar in model membranes and in bacteria \cite{gally81,scherer87,seelig90}, \\
3) headgroup structures are similar in PC, PE and PG lipids, while headgroup is more rigid in PS lipids \cite{wohlgemuth80,buldt81}. \\
Extensive discussion about structural details of PE, PG or PS headgroups do not exists (as far as I know), 
In contrast to PC lipids (see \cite{botan15} and references therein).

@@

Several simulations containing PE, PG and PS lipids have been published \cite{??}, \todo{List should be completed}
however, glycerol backbone and headgroup order parameters are not compared to
the experiments (based on superficial reading of literature).

Comparison between experiments and simulations with CHARMM GUI
parameters for PS lipids in Fig. \ref{HGorderParametersPOPS} suggest that
the model overestimates magnitudes in $\beta$ and g$_3$ order parameters
and forking in $\alpha$.
\begin{figure}[]
  \centering
  \includegraphics[width=9.0cm]{../Figs/HGorderparametersPOPS.eps}
  \caption{\label{HGorderParametersPOPS}
    Order parameters for POPS headgroup and glycerol backbone from simulations and experiments without CaCl$_2$ 
    (DOPS from \cite{browning80} contains 0.1M of NaCl, POPC:POPS mixture from \cite{roux90}).
    Absolute values are shown for experimental data, because signs are not known.
    Simulations values are -S$_{CH}$
  }
  \todo{More simulation data for lipids with different headgroups to be collected}
  \todo{CHARMM GUI simulation contains only counter ions as potassium.
    All experiments here contain some amount of sodium salt. The best
    ion concentrations for comparison should be figured out.}
  \todo{Experimental signs of the order parameters would highly useful.}
\end{figure}

Comparison between experiments and simulations with CHARMM GUI
parameters for PG lipids in Fig. \ref{HGorderParametersPOPG} suggest that
the model overestimates forking in $\gamma$ and $\alpha$ order parameters,
and magnitudes of $\beta$ and $g_1$ order parameters.
\begin{figure}[]
  \centering
  \includegraphics[width=9.0cm]{../Figs/HGorderparametersPOPG.eps}
  \caption{\label{HGorderParametersPOPG}
    Order parameters for PG headgroup and glycerol backbone from simulations and experiments without CaCl$_2$ 
    (POPG from \cite{borle85} contains 10mM of PIPES, DPPG from \cite{wohlgemuth80} contains 10mM PIPES and 100mM CaCl,
    E.Coli PG results from \cite{gally81}).
    Absolute values are shown, because signs are not known for experimental data.
    Glycerol backbone order parameters are negative in simulations, while $alpha$, $\beta$ and
    smaller $\gamma$ are positive.
  }
  \todo{More simulation data for lipids with different headgroups to be collected}
  \todo{CHARMM GUI simulation contains only counter ions as potassium.
    All experiments here contain some amount of sodium salt. The best
    ion concentrations for comparison should be figured out.}
  \todo{Experimental signs of the order parameters would highly useful.}
\end{figure}

@@

PC lipid headgroup order parameters can used to measure ion binding
affinity, because their magnitude is linearly proportional
to the amount of bound charge in bilayer \cite{seelig87,catte16}.
This molecular electrometer concept can be used also
for bilayers containing PC lipids mixed with charged lipids \cite{borle85,macdonald87,roux90}.
This is demonstrated in Figs \ref{OrderParametersWithCaCl},
\ref{OrderParametersWithCaClBELOW1M} and \ref{OrderParameterCHANGESWithCaClBELOW1M},
showing order parameters for PC headgroup $\alpha$ and $\beta$ carbons
as a function of CaCl$_2$ concentration in the presence of different amounts of
negatively charged PS or PG lipids.
\begin{figure}[]
  \centering
  \includegraphics[width=9.0cm]{../Figs/LIPIDSwithCaCl.eps}
  \caption{\label{OrderParametersWithCaCl}
    PC headgroup order parameters as a function of CaCl concentration from experiments containing charged lipids.
    Pure DPPC data from \cite{akutsu81}, pure POPC data from \cite{altenbach84}, 
    POPC:POPS mixture data from \cite{roux90}, POPC:POPG mixture data with 0.1M NaCl from \cite{macdonald87}
    and POPC:POPG mixture data without NaCl from \cite{borle85}.
  }
  \todo{Check the NaCl concentrations in the samples.}
\end{figure}
\begin{figure}[]
  \centering
  \includegraphics[width=9.0cm]{../Figs/LIPIDSwithCaClBELOW1M.eps}
  \caption{\label{OrderParametersWithCaClBELOW1M}
    Figure \ref{OrderParametersWithCaCl} zoomed to smaller concentrations.
  }
\end{figure}
\begin{figure}[]
  \centering
  \includegraphics[width=9.0cm]{../Figs/CHANGESwithCaCl.eps}
  \caption{\label{OrderParameterCHANGESWithCaClBELOW1M}
    The change of PC headgroup order parameters
    in the presence of different amount of negatively charged lipids
    respect to the values without added CaCl$_2$.
    The original data is the same as in Fig. \ref{OrderParametersWithCaCl}. 
  }
\end{figure}

PC headgroup order parameters increase when negatively charged
PS or PG are added to PC bilayer in the absense of added CaCl$_2$,
as expected based on electrometer concept \cite{seelig87}
(see Fig. \ref{OrderParametersWithCaClBELOW1M}).
%In electrometer concept this is explained by the tilting of
%headgroup more parallel to membrane normal \cite{??}.
Further, the order parameters decrease with the addition
of CaCl$_2$ and the decrease is more pronounced for systems with more
negatively charged lipids (see Fig. \ref{OrderParameterCHANGESWithCaClBELOW1M}).
At CaCl$_2$ concentrations ($\sim$ 50-300mM) where order parameters reach the values for pure PC,
the Ca2+ binding is presumably fully cancels the charge from negative lipids and
overcharging occurs above these concenterations.
The interpretation of this data and some other results has been that \cite{seelig90}
\begin{displayquote}
  {\it ''(i) Ca$^{2+}$ binds to neutral lipids (phosphatidylcholine, phosphatidylethanolamine) and negatively charged lipids
  (phosphatidylglycerol) with approximately the same binding constant of K = 10-20 M$^{-1}$; (ii) the free Ca$^{2+}$
  concentration at the membrane interface is distinctly enhanced if the membrane carries a negative surface
  charge, either due to protein or to lipid; (iii) increased inter-facial Ca$^{2+}$ also means increased amounts
  of bound Ca$^{2+}$ at neutral and charged lipids; (iv) the actual binding step can be described by a Langmuir
  adsorption isotherm with a 1 lipid:1 Ca$^{2+}$ stoichiometry, provided the interfacial concentration C$_M$, is
  used to describe the chemical binding equilibrium.''}
\end{displayquote}

Also the experimental order parameters for PS and PG headgroups
as a function of CaCl$_2$ concentration are shown in Fig. \ref{PSPGchangesWITHCaCl}.
\todo{These should be compared to simulations for potential structural interpretation of the changes.}
\begin{figure}[]
  \centering
  \includegraphics[width=9.0cm]{../Figs/PSPGwithCaCl.eps}
  \caption{\label{PSPGchangesWITHCaCl}
    PG and PS order parameters a function CaCl$_2$ concentration taken from \cite{borle85} and \cite{roux90}, respectively.
  }
  \todo{Get the small concentration data from the inserts}
\end{figure}
